\begin{abstract}
The past decade has exposed, with uncomfortable regularity, how vulnerable
modern supply chains are to disruption. Pandemic-induced shocks, geopolitical
instability, and the cumulative fragility of lean, globally distributed value
networks have pushed risk management to the centre of operations research
and managerial practice alike. In response, artificial intelligence has attracted
growing attention as a means of shifting the field away from reactive
contingency planning and toward more anticipatory, data-driven approaches
to resilience. Yet the body of work at this intersection remains scattered
across methodological traditions, risk taxonomies, and industrial sectors,
making it difficult to draw coherent conclusions about where progress has
genuinely been made and where significant gaps persist. This paper addresses 
that dispersion through a systematic literature review
conducted in accordance with the PRISMA~2020 guidelines. Drawing on searches
of the IEEE~Xplore and Web of Science databases, and covering publications
from 2019 to 2026, the review applies a rigorous two-stage screening protocol
to an initial corpus of 192 records, retaining 19 studies for qualitative
analysis. The analysis is organised around three research questions: how AI
is applied to mitigate supply chain risks, which risk categories are addressed,
and which techniques are deployed. The aim is not to advocate for any
particular approach but to offer a structured, evidence-based account of the
current state of the field and the directions most warranted by what the
literature, taken as a whole, reveals.

\keywords{Artificial Intelligence \and Supply Chain \and Risk Mitigation \and Machine Learning}
\end{abstract}
