\section{Introduction}

The past fifteen years have delivered a sequence of disruptions severe enough
to force a rethinking of assumptions that had gone largely unquestioned for
decades. The 2011 Tohoku earthquake brought automotive and electronics
production to a standstill across multiple continents, tracing supply
dependencies that few organisations had mapped in full. The COVID-19
pandemic, arriving nearly a decade later, did something similar but at a
scale and speed that left little time for adaptation: single-source strategies
that had looked efficient suddenly looked reckless, and shortages accumulated
faster than contingency plans could address them. The 2021 blockage of the
Suez Canal added a different kind of lesson, that a single point of
congestion, resolved within days, could translate into delays for hundreds of
organisations simultaneously. None of these events were, in hindsight,
unforeseeable in category. What they exposed was that the dominant logic of
supply chain design, built around leanness, cost optimisation, and the
assumption of reasonable stability, had quietly accumulated fragilities that
traditional management tools were not built to detect.

The response in both practice and research has been a turn toward
resilience, not as a vague aspiration but as a design property, one that
requires supply chains to anticipate disruptions rather than simply absorb
them. This shift has brought renewed attention to the question of what kinds
of tools can actually support that anticipatory capacity. Rule-based systems
and classical statistical models were built for environments with stable
historical distributions; they tend to underperform precisely when conditions
depart most sharply from past experience. Artificial intelligence offers a
different kind of capability. Machine learning and deep learning architectures
can identify non-linear patterns in high-dimensional, heterogeneous data that
would escape conventional analysis. Natural language processing can extract
disruption signals from news feeds, regulatory announcements, and social
media well before those signals appear in operational data. Graph-based models
can represent the relational structure of multi-tier supplier networks and
trace how failures propagate through them, something that tabular approaches
cannot do without losing the topology that makes propagation possible in the
first place.

These capabilities have attracted considerable research interest, and the
empirical literature has grown accordingly. The difficulty is that it has grown
in several directions at once, across methodological traditions, risk
taxonomies, industrial sectors, and geographic contexts that do not always
speak to one another. Identifying what the field actually knows, as opposed
to what individual studies demonstrate under particular conditions, requires
a more systematic effort of synthesis than the primary literature alone can
provide.

This paper undertakes that synthesis through a systematic literature review
conducted in accordance with the PRISMA~2020 framework. Starting from 192
records retrieved from IEEE~Xplore and Web of Science, and covering
publications from 2019 to 2026, the review applies a two-stage screening
protocol to arrive at 19 studies retained for qualitative analysis. Three
research questions organise the work: (RQ1)~how AI is applied to mitigate
supply chain risks; (RQ2)~which risk types are addressed; and (RQ3)~which AI
techniques are employed. The analysis goes beyond descriptive mapping to
identify structural gaps, geographic concentration in Asian research
contexts, limited coverage of cybersecurity and geopolitical risks, near-
absent integration of real-time data infrastructures, that point toward the
directions where the field most needs to develop.

The remainder of this paper is organised as follows. Section~2 establishes
the theoretical background. Section~3 describes the methodology.
Sections~4 and~5 present the selection results and content analysis.
Section~6 discusses key findings, emerging trends, and identified gaps.
Section~7 concludes.